A Wait-free Queue with Polylogarithmic Step Complexity

### 2023-01-09 podc_paper/algorithm-2.tex
minor edits

To determine the oldest block in a node $v$ to keep, the \opemph{SplitBlock} routine  first
finds the most recent block $B_{root}$ in the root that contains either an enqueue that has been dequeued
or a null dequeue using the \var{last} array.  All enqueues before this block are either
dequeued or will be dequeued by a dequeue that is in progress.  Once those pending dequeues have been
helped to complete, it is safe to discard any blocks in the root older than $B_{root}$, as well
as their subblocks.
The \op{SplitBlock} uses the \eleft\ and \eright\ fields to find the last subblock in $v$
of $B_{root}$ (or any older block in the root).
As \op{SplitBlock} is in progress, it is possible that some block that it needs in a node $v'$
along the path from $v$ to the root is discarded by another GC phase.
In this case, \op{SplitBlock} uses the last subblock in $v$ of the oldest block in $v'$ instead
(since $v'$ determined that all blocks older than that are safe to discard).

@@

\gc{
\Function{Block}{SplitBlock}{\typ{Node} v}
	\State \linecomment{figure out where to split $v$'s RBT}
	\If{$v=\var{root}$}
		\State $m \assign 0$
		\For{$k\assign 1..p$} $m\assign \max(m, v.\fld{last}[k])$
		\EndFor
		\State $B\assign \var{root}.\fld{blocks}[m-1]$
	\Else
		\State $B_p \assign \Call{SplitBlock}{v.parent}$
		\State \var{dir} \assign ($v=v.\fld{parent.left}$ ? \fld{left} : \fld{right})
		\State $B \assign v.\fld{blocks}[B_p.\edir]$
	\EndIf
	\State \linecomment If $B$ was discarded, use leftmost block instead
	\State \Return ($B = \nl$ ? $\Call{MinBlock}{v.\var{blocks}}$ : $B$)
\EndFunction{SplitBlock}
}

@@

\Function{boolean}{Refresh}{\typ{Node} \var{v}}
	\State \linecomment{try to append a new block to $v.\fld{blocks}$}
	\gc{
	\State $T \assign v.\fld{blocks}$\label{refreshReadTGC}
    \State $\var{h} \assign \Call{MaxBlock}{T}.\fld{index}+1$ \label{refreshReadMax}
    }
    \State \var{new} \assign\ \Call{CreateBlock}{\var{v, h}} \label{invokeCreateBlockGC}
    \If{\var{new = \nl}} \Return{\tr} \label{addOPGC}
    \Else  
    	\gc{
	    \State $T' \assign \Call{AddBlock}{v, T, \var{new}}$\label{refreshABGC}
		\State \Return \Call{CAS}{\var{v}.\fld{blocks}, $T$, $T'$}\label{casGC}
		}
    \EndIf

>>>>

To determine the oldest block in a node \var{v} to keep, the \opemph{SplitBlock} routine  first
finds the most recent block $B_{root}$ in the root that contains either an enqueue that has been dequeued
or a null dequeue using the \var{last} array.  All enqueues before this block are either
dequeued or will be dequeued by a dequeue that is in progress.  Once those pending dequeues have been
helped to complete, it is safe to discard any blocks in the root older than $B_{root}$, as well
as their subblocks.
The \op{SplitBlock} uses the \eleft\ and \eright\ fields to find the last subblock in \var{v}
of $B_{root}$ (or any older block in the root).
As \op{SplitBlock} is in progress, it is possible that some block that it needs in a node $\var{v}'$
along the path from \var{v} to the root is discarded by another GC phase.
In this case, \op{SplitBlock} uses the last subblock in \var{v} of the oldest block in $\var{v}'$ instead
(since $\var{v}'$ determined that all blocks older than that are safe to discard).

@@

\gc{
\Function{Block}{SplitBlock}{\typ{Node} v}
	\State \linecomment{figure out where to split \var{v}'s RBT}
	\If{$\var{v}=\var{root}$}
		\State $m \assign 0$
		\For{$k\assign 1..p$} $m\assign \max(m, \var{v}.\fld{last}[k])$
		\EndFor
		\State $B\assign \var{root}.\fld{blocks}[m-1]$
	\Else
		\State $B_p \assign \Call{SplitBlock}{\var{v}.\fld{parent}}$
		\State \var{dir} \assign ($\var{v}=\var{v}.\fld{parent.left}$ ? \fld{left} : \fld{right})
		\State $B \assign \var{v}.\fld{blocks}[B_p.\edir]$
	\EndIf
	\State \linecomment If $B$ was discarded, use leftmost block instead
	\State \Return ($B = \nl$ ? $\Call{MinBlock}{\var{v}.\var{blocks}}$ : $B$)
\EndFunction{SplitBlock}
}

@@

\Function{boolean}{Refresh}{\typ{Node} \var{v}}
	\State \linecomment{try to append a new block to $\var{v}.\fld{blocks}$}
	\gc{
	\State $T \assign \var{v}.\fld{blocks}$\label{refreshReadTGC}
    \State $\var{h} \assign \Call{MaxBlock}{T}.\fld{index}+1$ \label{refreshReadMax}
    }
    \State \var{new} \assign\ \Call{CreateBlock}{\var{v, h}} \label{invokeCreateBlockGC}
    \If{\var{new = \nl}} \Return{\tr} \label{addOPGC}
    \Else  
    	\gc{
	    \State $T' \assign \Call{AddBlock}{v, T, \var{new}}$\label{refreshABGC}
		\State \Return \Call{CAS}{\var{v}.\fld{blocks}, $T$, $T'$}\label{casGC}
		}
    \EndIf

### 2023-01-09 podc_paper/analysis.tex
minor edits; changed tournament tree to ordering tree
minor edits

\begin{lemma}\label{enqDeqTime}
Each \op{Enqueue} and null Dequeue takes $O(\log p)$ steps 
\Eric{In the thesis, Lemma 52 said enqueues take $O(\log^2 p)$ steps.  Was this a typo?}
and each non-null \op{Dequeue} takes
$O(\log p\log c + \log q_e+ \log q_d)$ steps,
where $q_d$ is the size of the queue when the \nf{Dequeue} is linearized and 
$q_e$ is the size of the queue when the \op{Enqueue} of the value returned is linearized.
\end{lemma}
\begin{proof}
An \op{Enqueue} or null \op{Dequeue} creates a block, appends it to the process's 
leaf and propagates it to the root.  The \op{Propagate}  does $O(1)$ steps 
at each node on the path from the process's leaf to the root.
A null \op{Dequeue} additionally calls \op{IndexDequeue}, which also does $O(1)$ steps
at each node on this path. 
So, the total number of steps for either type of operation is $O(\log p)$.

A non-null \op{Dequeue} must also search at line \ref{FRb} and call \op{GetEnqueue}
at line \ref{findAnswer}.
\Eric{Modified:}
By \Cref{dSearchTime}, the doubling search takes $O(\log(q_e+q_d+p))$ steps,
since the size of the queue can change by at most $p$ within one block (by \Cref{blockSize}).
\op{GetEnqueue} does a binary search within each node on a path from the root to a leaf.
Each node $v$'s search is within the subblocks of one block in $v$'s parent.
By \Cref{blockSize}, each search takes $O(\log c)$ steps, for a total of $O(\log p\log c)$ steps.
\end{proof}

\Eric{I left out the amortized bound, since the worst-case bound seems stronger}

>>>>

\begin{lemma}\label{enqDeqTime}
Each \op{Enqueue} and null Dequeue takes $O(\log p)$ steps 
and each non-null \op{Dequeue} takes
$O(\log p\log c + \log q_e+ \log q_d)$ steps,
where $q_d$ is the size of the queue when the \nf{Dequeue} is linearized and 
$q_e$ is the size of the queue when the \op{Enqueue} of the value returned is linearized.
\end{lemma}
\begin{proof}
An \op{Enqueue} or null \op{Dequeue} creates a block, appends it to the process's 
leaf and propagates it to the root.  The \op{Propagate}  does $O(1)$ steps 
at each node on the path from the process's leaf to the root.
A null \op{Dequeue} additionally calls \op{IndexDequeue}, which also does $O(1)$ steps
at each node on this path. 
So, the total number of steps for either type of operation is $O(\log p)$.

A non-null \op{Dequeue} must also search at line \ref{FRb} and call \op{GetEnqueue}
at line \ref{findAnswer}.
By \Cref{dSearchTime}, the doubling search takes $O(\log(q_e+q_d+p))$ steps,
since the size of the queue can change by at most $p$ within one block (by \Cref{blockSize}).
\op{GetEnqueue} does a binary search within each node on a path from the root to a leaf.
Each node $v$'s search is within the subblocks of one block in $v$'s parent.
By \Cref{blockSize}, each search takes $O(\log c)$ steps, for a total of $O(\log p\log c)$ steps.
\end{proof}
